---
title: "Group 22 Analysis"
author: "Group 22"
number-sections: true
format: 
  html:
    embed-resources: true
editor_options:
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false

# the lines above ensure that the R code in this block (along with any warning messages) will not be shown
# load any libraries you will use here. E.g.
library(tidyverse)
library(skimr)
library(janitor)
library(here)

```
# Introduction {#sec-intro}

Email spam detection is an important application of statistical modelling because unwanted messages can reduce communication efficiency and increase security risks.  
For Group 22, the dataset contains several text-related variables that measure the frequency of specific characters or words in an email, together with a binary response variable indicating whether the email is classified as spam.

The aim of this analysis is to investigate which explanatory variables are associated with spam classification, and to prepare an appropriate basis for later generalised linear modelling.  
Before fitting any formal model, it is necessary to import the dataset, inspect its structure, and carry out basic cleaning checks to ensure that the analysis is reproducible and well documented.

# Exploratory Analysis {#sec-EA}
## Data import

This section begins by importing the original dataset for Group 22 from the project directory.
```{r}
# Read the original dataset from the project folder
spam_raw <- read_csv(here("dataset22.csv"))
```


## Initial inspection

The dataset is then inspected to understand its structure, dimensions, variable names, and the first few observations.
```{r}
# Display the first few rows
head(spam_raw)

# Check the structure of the dataset
glimpse(spam_raw)

# Check the dimensions of the dataset
dim(spam_raw)

# Show variable names
names(spam_raw)
```
The initial inspection shows that the dataset contains 920 observations and 7 variables.  
The response variable `yesno` is initially stored as a character variable, while the remaining variables are numeric.

## Data cleaning

The response variable is converted to the correct type, and basic checks are carried out for missing values and duplicated rows.
```{r}
# Create a cleaned copy for further processing
spam_clean <- spam_raw

# Convert the response variable to a factor
spam_clean <- spam_clean %>%
  mutate(yesno = factor(yesno))

# Check missing values in each column
colSums(is.na(spam_clean))

# Check duplicated rows
sum(duplicated(spam_clean))
```
The cleaning step confirms that there are no missing values in the dataset.  
The response variable `yesno` is converted to a factor so that it can be used appropriately in later classification modelling.  
In addition, 210 duplicated rows are detected. These are retained at this stage and noted for later discussion, rather than removed without justification.

## Summary of cleaned data

Finally, a short summary of the cleaned dataset is produced to confirm that it is ready for later modelling.
```{r}
# Produce a quick overall summary of the cleaned dataset
skim(spam_clean)

# Frequency table for the response variable
tabyl(spam_clean$yesno)
```
The summary output confirms that the cleaned dataset is ready for subsequent analysis.  
The response variable is moderately imbalanced, with 557 non-spam emails (60.5%) and 363 spam emails (39.5%).

# Formal Analysis {#sec-FA}

# Conclusions {#sec-con}
