---
title: "Group 22 Analysis"
author: "Group 22"
number-sections: true
format:
  html:
    embed-resources: true
editor_options:
  chunk_output_type: console
---

```{r}
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(skimr)
library(janitor)
library(here)
library(GGally)
```

# Introduction {#sec-intro}

Email spam detection is an important application of statistical modelling because unwanted messages can reduce communication efficiency and increase security risks.  
For Group 22, the dataset contains several text-related variables that measure the frequency of specific characters or words in an email, together with a binary response variable indicating whether the email is classified as spam.

The aim of this analysis is to investigate which explanatory variables are associated with spam classification, and to prepare an appropriate basis for later generalised linear modelling.  
Before fitting any formal model, it is necessary to import the dataset, inspect its structure, and carry out basic cleaning checks to ensure that the analysis is reproducible and well documented.

# Exploratory Analysis {#sec-EA}

## Data import

This section begins by importing the original dataset for Group 22 from the project directory.

```{r}
# Read the original dataset from the project folder
spam_raw <- read_csv(here("dataset22.csv"))
```

## Initial inspection

The dataset is then inspected to understand its structure, dimensions, variable names, and the first few observations.

```{r}
# Display the first few rows
head(spam_raw)

# Check the structure of the dataset
glimpse(spam_raw)

# Check the dimensions of the dataset
dim(spam_raw)

# Show variable names
names(spam_raw)
```

The initial inspection shows that the dataset contains 920 observations and 7 variables.  
The response variable `yesno` is initially stored as a character variable, while the remaining variables are numeric.

## Data cleaning

The response variable is converted to the correct type, and basic checks are carried out for missing values and duplicated rows.

```{r}
# Create a cleaned copy for further processing
spam_clean <- spam_raw

# Convert the response variable to a factor
spam_clean <- spam_clean %>%
  mutate(yesno = factor(yesno))

# Check missing values in each column
colSums(is.na(spam_clean))

# Check duplicated rows
sum(duplicated(spam_clean))
```

The cleaning step confirms that there are no missing values in the dataset.  
The response variable `yesno` is converted to a factor so that it can be used appropriately in later classification modelling.  
In addition, 210 duplicated rows are detected. These are retained at this stage and noted for later discussion, rather than removed without justification.

## Summary of cleaned data

Finally, a short summary of the cleaned dataset is produced to confirm that it is ready for later modelling.

```{r}
# Produce a quick overall summary of the cleaned dataset
skim(spam_clean)

# Frequency table for the response variable
tabyl(spam_clean$yesno)
```

The summary output confirms that the cleaned dataset is ready for subsequent analysis.  
The response variable is moderately imbalanced, with 557 non-spam emails (60.5\%) and 363 spam emails (39.5\%).

## Dataset Overview

We start with the response variable `yesno`, which tells us whether an email is spam (`y`) or not spam (`n`).

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-align: center
#| fig-width: 4.5
#| fig-height: 3.5
#| label: fig-barplot1
#| fig-cap: Distribution of Spam and Non-Spam Emails.

ggplot(spam_clean, aes(x = yesno, fill = yesno)) +
  geom_bar() +
  labs(
    title = "Distribution of Spam and Non-Spam Emails",
    x = "Email Type",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

As shown in @fig-barplot1, there are more non-spam emails than spam emails in the dataset. The difference is there, but it is not very large, so class imbalance does not look like a major issue at this stage.

## Distribution of Capital Letter Sequences

We then look at `crl.tot`, which measures the total length of uninterrupted capital letter sequences in each email.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-align: center
#| fig-width: 4.5
#| fig-height: 3.5
#| label: fig-histplot2
#| fig-cap: Distribution of Continuous Capital Letter Sequences.

ggplot(spam_clean, aes(x = crl.tot)) +
  geom_histogram(bins = 40) +
  labs(
    title = "Distribution of Continuous Capital Letter Sequences",
    x = "Capital Letter Sequence Length",
    y = "Frequency"
  ) +
  theme_minimal()
```

In @fig-histplot2, the distribution is clearly right-skewed. Most emails have fairly small values of `crl.tot`, while only a small number have much larger values. So long capital letter sequences are not common, but they do appear in some emails. This also suggests that a few observations may have a stronger influence than the rest.

## Dollar Sign Frequency and Spam

This plot compares the frequency of dollar signs in spam and non-spam emails. The variable `dollar` measures how often the dollar symbol appears as a percentage of the total number of characters in the email.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-align: center
#| fig-width: 4.5
#| fig-height: 3.5
#| label: fig-boxplot3
#| fig-cap: Dollar Sign Frequency by Email Type.

ggplot(spam_clean, aes(x = yesno, y = dollar, fill = yesno)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Dollar Sign Frequency by Email Type",
    x = "Email Type",
    y = "Dollar Sign Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Looking at @fig-boxplot3, spam emails seem to have higher values of `dollar` overall. Many non-spam emails are very close to zero, while the spam group is more spread out and includes some larger values. There is still overlap between the two groups, so the separation is not perfect, but `dollar` looks useful as a predictor.

## Exclamation Mark Frequency and Spam

We also compare the use of exclamation marks across the two email types. The variable `bang` gives the percentage of exclamation marks in each email.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-align: center
#| fig-width: 4.5
#| fig-height: 3.5
#| label: fig-boxplot4
#| fig-cap: Exclamation Mark Frequency by Email Type.

ggplot(spam_clean, aes(x = yesno, y = bang, fill = yesno)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Exclamation Mark Frequency by Email Type",
    x = "Email Type",
    y = "Exclamation Mark Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

@fig-boxplot4 shows a similar pattern, although the difference looks a little clearer here. Non-spam emails are again concentrated near zero, while spam emails tend to have higher values and a wider spread. Compared with `dollar`, `bang` seems to give slightly clearer separation between the two groups.

## Keyword Frequency: "money"

This figure looks at the use of the word `money` in spam and non-spam emails. The variable `money` measures how frequently the word appears relative to the total number of characters in an email.

```{r}
#| echo: false
#| warning: false
#| message: false
#| fig-align: center
#| fig-width: 4.5
#| fig-height: 3.5
#| label: fig-boxplot5
#| fig-cap: Frequency of the Word "Money" by Email Type.

ggplot(spam_clean, aes(x = yesno, y = money, fill = yesno)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Frequency of the Word 'Money' by Email Type",
    x = "Email Type",
    y = "Frequency of 'Money'"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

@fig-boxplot5 suggests that the word `money` appears more often in spam emails than in non-spam emails. Similar to the previous plots, many non-spam emails have values near zero, while spam emails show a wider spread and more high values. This is consistent with the idea that financially related language is more common in spam.

## Relationships Between Predictors

The next step is to examine the relationships between the explanatory variables. This helps us check whether any predictors are strongly related to each other.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-corplot6
#| fig-cap: Pairwise Relationships Between Explanatory Variables.

GGally::ggpairs(
  spam_clean[, c("crl.tot", "dollar", "bang", "money", "n000", "make")]
)
```

From @fig-corplot6, most of the predictors look only weakly related. The clearest relationship seems to be between `dollar` and `n000`, but even that does not look especially strong. Overall, the plot does not suggest a serious multicollinearity problem, so these variables can reasonably be taken forward into the formal model.

## Exploratory Analysis Summary

Overall, the exploratory analysis suggests that several text-based features differ between spam and non-spam emails, especially dollar signs, exclamation marks, and financial terms. At the same time, the relationships between predictors seem fairly modest. Taken together, these results give some support for using these variables in the later logistic regression analysis.

# Formal Analysis {#sec-FA}

# Conclusions {#sec-con}
